---
title: "Primary Visualizations"
author: "Neville Taraporevala"
date: "2023-04-20"
output: html_document
---

# Preliminary Visualizations

Here are some initial visualizations of our data!

## Maps

Here is a map of all the sightings broken down by species in Wichita.

```{r firstvis, eval = T, include = T, warning = FALSE, echo = F}
library(usmap)

sedgwick <- map_data("county", "kansas") %>% 
  select(lon = long, lat, group, id = subregion) %>% 
  filter(id == "sedgwick")

ggplot(sedgwick, aes(lon, lat, group = group)) +
  geom_polygon(fill = "white", colour = "grey50") +
  geom_point(data = filter(sightings_b_h_f, Species == "Coyote" | Species == "Red Fox"),
             aes(x = x_long, y = y_lat, group = 1, alpha = 0.3, color = Species))+
  theme_bw()+
  xlim(-98,-97)+ylim(37.3,38)+
  labs(title = "Canid Sightings in Sedgwick County, KS", x = "Longitude", y = "Latitude")
```

```{r importGgmap, eval = T, include = T, warning = F, message = F, echo = F}
library(ggmap)
library(osmdata)
ict_basemap <- get_map(getbb("Wichita"), zoom=11, maptype = 'terrain', source = 'stamen')
```

And using `ggmap` we can show it over a basemap.

```{r makemap, eval = T, include = T, warning = F, echo = F}
ggmap(ict_basemap)+
  geom_point(data = filter(sightings_b_h_f, Species == "Coyote" | Species == "Red Fox"), 
             aes(x = x_long, y = y_lat, group = 1, color = Species, alpha = 0.3))+
  theme_bw()
```

## How people feel seeing canids

And here are some people's feelings broken down by which animal they saw.

```{r feelingsgraphs, eval = T, include = T, echo = F}
ggplot(data = filter(sightings_b_h_f, Species == "Coyote" | Species == "Red Fox"),
       aes(x = Happy, fill = Species)) +
  geom_bar() +
  labs(x = "Not Happy                                               Happy  ",
       y = "Number of People",
       title = "People who are happy to see red foxes vs. coyotes") +
  theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

ggplot(data = filter(sightings_b_h_f, Species == "Coyote" | Species == "Red Fox"),
       aes(x = Worried, fill = Species)) +
  geom_bar() +
  labs(x = "Not Worried                                             Worried  ",
       y = "Number of People",
       title = "People who are worried to see red foxes vs. coyotes") +
  theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

```

As you can see, a high proportion of people are worried about seeing coyotes and are happy to see foxes.
